Designing Types for R, Empirically

### 2020-05-16 paper/main.tex
stuff

\paragraph{Matrices}

Matrices represent another layer of complexity on top of vectorized
primitives.  Internally, matrices are simply vectors with class ``matrix''
and a \code{dims} attribute indicating matrix dimensions, and while not
codified in the language semantics, many internal functions will coerce
vectors to matrices automatically.  Relying on automatic coercion is unwise,
and many functions check to ensure that their arguments are matrices.
The \code{rowWeightedMeans} function calculates the weighted means of rows
or columns of a matrix \code{x}.  This function exhibits good practice in
type checking arguments, rightly producing a message to the user indicating
that an explicit matrix should be passed. We captured this with our class
types, and indeed \typetracer inferred a good type for this function,
allowing \contractr to perform the check present in this code automatically,
notifying the users passing ill-typed arguments.

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
  rowWeightedMeans <- function(x, w=NULL, rows=NULL, cols=NULL, na.rm=FALSE, ...) {
    if (!is.matrix(x)) 
      .Defunct(msg = sprintf("Argument 'x' should be a matrix.)
\end{lstlisting}

Thus far we've seen instances where we introduce types that subsume existing
argument validation code, but our intention is not to subsume {\it all} such
code. Consider:

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
  rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL, na.rm = FALSE) {
    n <- ncol(x)
    if (length(w) != n) 
       stop("The length of 'w' does not match columns in 'x': ")
\end{lstlisting}

This code is from later in the same \code{rowWeightedMeans} function
pdiscussed previously.  Here, the length of the weights vector \code{w} is
being checked for equality with the number of columns in the matrix
\code{x}.  Specifying this relationship as a type amounts to {\it dependent
  typing}, which we believe too complex to retrofit onto a dynamic data
science language such as R.  Ultimately, we decided to leave these kinds of
checks up to the programmer.

\paragraph{Data Frames}

One of the most popular classes in R is the \code{data.frame} class, making up nearly 8\% of observed classes.  Data
frames and the derivative \code{tibble} (from the \code{tidyverse}
ecosystem) and \code{data.table} (from the \code{data.table} package) types
underpin a huge amount of data analysis in R.

One way to deal with data frames is through the struct type, with a named
field for each column of the data frame, but as mentioned previously structs
introduced undue noise into \typetracer's analysis results.  Further
complicating data frames is that many functions built to operate on them
operate in a name-agnostic way.  For instance, the \code{tidyverse} package
ecosystem allows programmers to pass column names to functions which operate
on their data frames.  In base R, idiomatic data frame use is to use string
column names to select rows from the frame (unless only a single column is
of interest, wherein the \code{$} syntax is appropriate).

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
data(cars)
# tidyverse: Filter rows where speed is > 20. 
fast_cars_tidy <- filter(cars, speed > 20)
# base R: Get all rows where the "speed" column has a value > 20.
fast_cars_base <- cars[cars[, "speed"] > 20, ] # base example
\end{lstlisting}

Doing justice to this type will be an endeavour in and of itself, requiring
reconciliation of base R's \code{data.frame}, \code{tidyverse}'s
\code{tibble}, and \code{data.table}'s \code{data.table}.  Developing useful
types for the \code{tidyverse} ecosystem may well involve

It will also require proper treatment of function types, and perhaps even
parametric types, which dynamic analyses such as \typetracer are unable to
discern soundly.

In the context of our type annotation language, we ascribe the type \CLASS{data.frame} to \code{data.frame}-typed values.
Consider the following code snippet, showing the type signature for \code{dplyr}'s \code{all_equal} function which checks two \CLASS{data.frame} for equality.
\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
#' @type <target: class<data.frame>, current: class<data.frame>, ignore_col_order: lgl, 
#'         ignore_row_order: lgl, convert: lgl, ...> => (chr | lgl)
dplyr::all_equal <- function(target, current, ignore_col_order = TRUE,
                      ignore_row_order = TRUE, convert = FALSE, ...) { ... }
\end{lstlisting}

>>>>

\paragraph{Matrices}
Matrices are instances of the eponymous class. They have a \code{dims}
attribute indicating dimensions, and while not codified in the language
semantics, many internal functions coerce vectors to matrices automatically.
The \code{rowWeightedMeans} function calculates the weighted means of rows.
The programmer added a type check for \code{x}.

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
  rowWeightedMeans <- function(x, w=NULL, rows=NULL, cols=NULL, na.rm=FALSE, ...) {
    if (!is.matrix(x))   .Defunct(msg = sprintf("'x' should be a matrix.)
\end{lstlisting}

\paragraph{Data Frames}
One of the most popular classes in R is the \code{data.frame} class, making
up nearly 8\% of observed classes.  Data frames and the derivative
\code{tibble} and \code{data.table} types underpin many popular uses of R.
One way to deal with data frames is through the struct type, with a named
field for each column of the data frame, but as mentioned previously structs
introduced undue noise.  Further complicating data frames is that many
functions built to operate on them operate in a name-agnostic way.  For
instance, the \code{tidyverse} package ecosystem allows programmers to pass
column names to functions which operate on their data frames.  In base R,
idiomatic data frame use is to use string column names to select rows from
the frame (unless only a single column is of interest, wherein the \code{$}
syntax is appropriate).
